Making Sequential Programs Deterministic

### 2019-04-05 java-design.tex
Move a figure
Edits in secton 3

\TheDeterminismChecker treats arrays analogously to collections:
the semantics and rules are similar even though the syntax varies.
For simplicity this
paper uses the term ``collection'' and the type \<Collection> to represent
collections and arrays.
Collections include any type (including user-defined classes) that
implements the \<Iterable> or \<Iterator> interfaces; this includes all
Java collections such as \<List>s and \<Set>s.
Our implementation also supports any type that implements the \<Map>
interface.

    // Client code
    public class MapUtils {
        public static <K extends @PolyDet Object, V extends @PolyDet Object> 
                                @Det String toString(@PolyDet Map<K,V> map) {
            ...
            for(Map. @Det Entry<K,V> entry : map.entrySet()) { ... }
        }
    }
    [ERROR] MapUtils.java:[20,50] [enhancedfor.type.incompatible] 
    incompatible types in enhanced for loop.
    found : @PolyDet Entry<K extends @PolyDet Object,V extends @PolyDet Object>
    required: @Det Entry<K extends @PolyDet Object,V extends @PolyDet Object>
    \end{verbatim}
    \caption{Example: Error detected by \theDeterminismChecker in \<scribe-java>~\cite{nondex}.}
%    \todo{Use real examples of bugs from case studies, throughout.  Artificial
%        examples are less compelling (and lead the reader to wonder whether no
%        real example exist).  Put the first example of an error in a figure in
%        Section 1, to help with motivating the problem.}
    \label{code-determinism}
\end{figure}

@@

A \<Map> is deterministic if its \<entrySet> is deterministic.
In other words, iterating over this \<entrySet> produces the same values in the same order across executions.
\Cref{code-determinism} shows that the determinism qualifier on the return type of \<entrySet()> is the
same as that on the receiver.  That is, its type (ignoring type arguments) is
$\forallt{\kappa} \kappa\ \<Map> \rightarrow \kappa\ \<Set>$\ .

We have not yet found any implementation of the \<Map>
interface in which \<entrySet()> \<keySet()>, and \<values()> behave
differently with respect to determinacy:  either all, or none, or deterministic.

@@

\TheDeterminismChecker prohibits the creation of a \codeid{@Det HashMap} or an \codeid{@OrderNonDet TreeMap}.
\Cref{fig-creation-rules} formalizes the type well-formedness rules for
Java \<Map>s and \<Set>s.

@@

The JDK implements \<LinkedHashMap> as a subclass of \<HashMap>
in order to inherit parts of \<HashMap>'s implementation.
%\todo{Talk about this design choice. Josh Bloch's book: "If the code surrounding the first declaration depended on LinkedHashSet's ordering policy, then it would be incorrect to substitute HashSet for LinkedHashSet in the declaration, because HashSet makes no guarantee concerning iteration order."} 
%\todo{The above is not clear.  What are the consequences for our
%  implementation.  What goes wrong or is hard to annotate or is surprising?
%  I'm not sure what is the point of this paragraph.}
\todo{From the discussion in this section, it is clear that \<LinkedHashMap> and \<HashMap> do not
have a subtyping relationship with respect to their determinism types. We looked at the implementation of
these two classes in the JDK and found that \<LinkedHashMap> doesn't reuse code that corresponds to 
iterating over its keys or values with \<HashMap>.}

@@

\begin{itemize}
\item
In \theDeterminismCheckerImplementation,
type polymorphism is handled by Java's generics mechanism, which
\theDeterminismChecker fully supports, including class and method generics,
inference, etc.
Given the Java declaration
\begin{Verbatim}
<T> T identity(T arg) { return arg; }
\end{Verbatim}
the type of \<identity> is $\forallt{\tau} \tau \rightarrow \tau$, and
the type of \<identity(arg)> is the same as the type of
\<arg>.  A programmer can also write a type qualifier on a type argument,
as in \codeid{@NonDet List<@NonDet Integer>}.
\item
Basetype polymorphism is enabled by writing a type qualifier on a use of a
type variable, which overrides the type qualifier at the instantiation
site.
For example, a \<choose> operation on sets could be defined in Java as
\begin{Verbatim}
interface ChooseableSet<T> implements Set<T> {
  // return an arbitrary element from the set
  @NonDet T choose();
}
\end{Verbatim}
\item
Java does not provide a syntax that can be used for qualifier polymorphism,
so \theDeterminismChecker uses a special type qualifier name, \<@PolyDet>.  (\<@PolyDet> stands
for ``polymorphic determinism qualifier.)
A qualifier-polymorphic method \<m> with signature $\forallt{\kappa} \kappa\ \<int> \times \<Det boolean> \rightarrow
\kappa\ \<String>$ would be written in our dialect of Java as
\begin{Verbatim}
@PolyDet String m(@PolyDet int, @Det boolean)
\end{Verbatim}
Each use of \<@PolyDet> stands for a use of the qualifier variable
$\kappa$, and there is no need to declare the qualifier variable $\kappa$.
\TheDeterminismChecker currently supports
qualifier
polymorphism on methods but \emph{not} on classes.
Therefore, \<@PolyDet> may be written in methods (signatures and bodies)
but not on fields.
\end{itemize}

Qualifier  polymorphism is common on methods that a programmer might think
of as deterministic.  For example, an addition method could be defined as

@@

This can be used in more contexts than

@@

@@

\TheDeterminismChecker applies a default qualifier at each unqualified Java
basetype (except uses of type variables, which already stand for a type that was
defaulted at the instantiation site).
This does not change the expressivity of the type system; it merely makes
the system more convenient to use by reducing programmer effort and code clutter.
Defaulted type qualifiers are not trusted:  they are type-checked just as
explicitly-written ones are.

@@

\noindent
This choice type-checks if the method body does not make calls to any
interfaces that require \<@Det> arguments or produce \<@NonDet> results.
Otherwise, the programmer must write an explicit type qualifier.

@@

As an exception to the above rules about return types, if a method's formal
parameters (including the receiver) are all \<@Det> --- that is, there are
no unannotated or \<@PolyDet> formal parameters --- then an unannotated
return type defaults to \<@Det>.  This is particularly useful for methods
that take no formal parameters.  A type like $\forallt{\kappa} ()
\rightarrow \kappa int$ does not make sense, becauses there is no way to
instantiate the type argument $\kappa$.  Treating the type as $()
\rightarrow \<Det> int$ permits just as many uses.

The default annotation for local variables is \<@NonDet>, which is the top
qualifier in the hierarchy.  \Cref{sec:dataflow-java} explains why.

@@

\todo{Relate this to defaulting.}

It also performs dataflow analysis and refines the types of these locals, thereby reducing the qualifier burden 
on the programmer. Consider the example below:
\begin{verbatim}
void foo(@Det int x) {
    @NonDet int y = x;    // Type of y gets refined to @Det.
    @Det int z = y;       // No error here.
}
\end{verbatim}
In the example above, even though the local variable \codeid{y} was declared to be \codeid{NonDet}, it gets
type refined to \codeid{Det} after the first assignment statement. As the result, the assignment of \codeid{y}
to a \codeid{Det} variable \codeid{z} is sound. In the following example,
\begin{verbatim}
void foo(@NonDet Object x, @Det Object y) {
    x = y;                // No type refinement for x.
    @Det Object z = x;    // Error.
}
\end{verbatim}
\theDeterminismChecker reports a warning at the second assignment statement of \codeid{foo}. Type refinement only applies to locals. Refining non-locals can result in side effects making the analysis unsound.

With the above rules, \theDeterminismChecker will automatically annotate a local collection with type \<NonDet> and its
type arguments will get the type qualifier \<Det>. For example, a locally declared \codeid{List} of \codeid{Strings} will be automatically annotated
as \codeid{NonDet List<Det String>}. Recall from~\cref{collection-rules} that \theDeterminismChecker does not allow this type on
any collection. While this is still sound, it will result in \theDeterminismChecker reporting a lot of errors because every locally declared 
collection gets annotated with this invalid type. To avoid this situation, we make an exception to the defaulting rules for local collections
and annotate their type parameters also with \<NonDet>. So a local list of Strings gets the default qualifier of \codeid{NonDet List<NonDet String>}. While this defaulting eliminates the possibility of automatically annotating locals collections with an invalid
type, it could result in a high number of false positives. In the following code snippet,
\begin{verbatim}
void test(@Det List<@Det String> argList) {
    List<String> localList = argList;
}    
\end{verbatim}
\theDeterminismChecker would report an error at the assignment statement since the inferred type for \codeid{localList} is 
\codeid{NonDet List<NonDet> String} and since collections types are invariant.
We therefore recommend that programmers explicitly annotate
all local variables that are collections to reduce false positives.

\subsection{Precision improvements}\label{sec:precision-java}

>>>>

\TheDeterminismChecker treats arrays analogously to collections:
the semantics and rules are similar even though the syntax varies.
For simplicity this
paper uses the term ``collection'' and the type \<Collection> to represent
collections and arrays.
Collections include any type (including user-defined classes) that
implements the \<Iterable> or \<Iterator> interfaces; this includes all
Java collections such as \<List>s and \<Set>s.

@@

A \<Map> is deterministic if its \<entrySet> is deterministic.
In other words, iterating over this \<entrySet> produces the same values in the same order across executions.
The determinism qualifier on the return type of \<entrySet()> is the
same as that on the receiver (\cref{code-determinism}).
That is, its type (ignoring type arguments) is
$\forallt{\kappa} \kappa\ \<Map> \rightarrow \kappa\ \<Set>$\ .

We have not yet found any implementation of the \<Map>
interface in which \<entrySet()> \<keySet()>, and \<values()> behave
differently with respect to determinacy:  either all three methods, or none, are deterministic.

@@

\noindent
\TheDeterminismChecker prohibits the creation of a \codeid{@Det HashMap} or an \codeid{@OrderNonDet TreeMap}.
\Cref{fig-creation-rules} formalizes the type well-formedness rules for
Java \<Map>s and \<Set>s.

@@

@@

\begin{itemize}
\item
In \theDeterminismCheckerImplementation,
type polymorphism is handled by Java's generics mechanism, which
\theDeterminismChecker fully supports, including class and method generics,
inference, etc.
Given the Java declaration
\begin{Verbatim}
<T> T identity(T param) { return param; }
\end{Verbatim}
the type of \<identity> is $\forallt{\tau} \tau \rightarrow \tau$, and
the type of \<identity(x)> is the same as the type of
\<x>.  A programmer can also write a type qualifier on a type argument,
as in \codeid{@NonDet List<@NonDet Integer>}.
\item
Basetype polymorphism is enabled by writing a type qualifier on a use of a
type variable, which overrides the type qualifier at the instantiation
site.
For example, a \<choose> operation on sets could be defined in Java as
\begin{Verbatim}
interface ChooseableSet<T> implements Set<T> {
  // return an arbitrary element from the set
  @NonDet T choose();
}
\end{Verbatim}
\item
Java does not provide a syntax that can be used for qualifier polymorphism,
so \theDeterminismChecker uses a special type qualifier name, \<@PolyDet>.  (\<@PolyDet> stands
for ``polymorphic determinism qualifier''.)
A qualifier-polymorphic method \<m> with signature
$\forallt{\kappa} \kappa\ \<int> \times \<Det boolean>
\rightarrow \kappa\ \<String>$ would be written as
\begin{Verbatim}
@PolyDet String m(@PolyDet int, @Det boolean)
\end{Verbatim}
Each use of \<@PolyDet> stands for a use of the qualifier variable
$\kappa$, and there is no need to declare the qualifier variable $\kappa$.
\TheDeterminismChecker currently supports
qualifier
polymorphism on methods but \emph{not} on classes.
Therefore, \<@PolyDet> may be written in methods (signatures and bodies)
but not on fields.
\end{itemize}

Qualifier  polymorphism is common on methods that a programmer might think
of as deterministic.  For example, an addition method should be defined as

@@

\noindent
This can be used in more contexts than

@@

All of our syntax is legal Java code that can be compiled with any Java 8
or later compiler.

    // Client code
    public class MapUtils {
        public static <K extends @PolyDet Object, V extends @PolyDet Object> 
                                @Det String toString(@PolyDet Map<K,V> map) {
            ...
            for (Map. @Det Entry<K,V> entry : map.entrySet()) { ... }
        }
    }
    [ERROR] MapUtils.java:[20,50] [enhancedfor.type.incompatible] 
    incompatible types in enhanced for loop.
    found : @PolyDet Entry<K extends @PolyDet Object,V extends @PolyDet Object>
    required: @Det Entry<K extends @PolyDet Object,V extends @PolyDet Object>
    \end{verbatim}
    \caption{Example: Error detected by \theDeterminismChecker in \<scribe-java>~\cite{nondex}.}
%    \todo{Use real examples of bugs from case studies, throughout.  Artificial
%        examples are less compelling (and lead the reader to wonder whether no
%        real example exist).  Put the first example of an error in a figure in
%        Section 1, to help with motivating the problem.}
    \label{code-determinism}
\end{figure}

@@

\TheDeterminismChecker applies a default qualifier at each unqualified Java
basetype (except uses of type parameters, which already stand for a type that was
defaulted at the instantiation site where a type argument was supplied).
This does not change the expressivity of the type system; it merely makes
the system more convenient to use by reducing programmer effort and code clutter.
Defaulted type qualifiers are not trusted:  they are type-checked just as
explicitly-written ones are.

@@

\noindent
This choice type-checks if the method body does not make calls to any
interfaces that require \<@Det> arguments or produce \<@NonDet> results.
Otherwise, the programmer must write explicit type qualifiers in the method
signature.

@@

As an exception to the above rules about return types, if a method's formal
parameters (including the receiver) are all \<@Det> --- that is, there are
no unannotated or \<@PolyDet> formal parameters --- then an unannotated
return type defaults to \<@Det>.  This is particularly useful for methods
that take no formal parameters.  A type like $\forallt{\kappa} ()
\rightarrow \kappa\ \<int>$ does not make sense, because there is no basis
on which to choose an
instantiation for the type argument $\kappa$.  Treating the type as $()
\rightarrow \<Det>\ \<int>$ permits just as many uses.

Types are inferred for unannotated local variables; see \cref{sec:dataflow-java}.

@@

\OurTypeSystem is
flow-sensitive~\cite{Hunt:2006:FST:1111037.1111045,Adams:2011:FTR:2048066.2048105,Sui:2016:OSU:2950290.2950296}.
That is, an expression may have
a different type on every line of the program, based on assignments and
side effects.  The current type must always be consistent with the declared
type, if any.
Consider the example below:

\begin{verbatim}
  @NonDet int x = ...;
                          // The type of x is @NonDet int
  x = 42;
                          // The type of x is @Det int
  @Det int y = x;
  x = random();
                          // The type of x is @NonDet int
  y = random();           // Error: y is declared as @Det int
\end{verbatim}

Arbitrary expressions can be given more specific types, including fields
and pure method calls.  A type refinement is lost whenever a side effect
might affect the value.  For example, type refinements to all fields are
lost whenever a non-pure method is called.

This flow-sensitive type refinement achieves local variable inference,
freeing progammers from writing many local variable types.

\todo{Mike doesn't understand the below.  An example would help clarify it.}

In the example above, even though the local variable \codeid{y} was declared to be \codeid{NonDet}, it gets
type refined to \codeid{Det} after the first assignment statement. As the result, the assignment of \codeid{y}
to a \codeid{Det} variable \codeid{z} is sound. In the following example,
\begin{verbatim}
void foo(@NonDet Object x, @Det Object y) {
    x = y;                // No type refinement for x.
    @Det Object z = x;    // Error.
}
\end{verbatim}
\theDeterminismChecker reports a warning at the second assignment statement of \codeid{foo}. Type refinement only applies to locals. Refining non-locals can result in side effects making the analysis unsound.

\subsection{The environment}\label{sec:environment-java}
